---
title: "Behavioral Data Preprocessing: Variable Recoding"
output: html_notebook
---

<!--
Purpose:      Load and recode raw HCP-Aging behavioral/demographic variables for use in
              loneliness brain network analyses.

Inputs:
  - ndar_subject01.csv           NDA demographics package;
                                 contains age, sex, race, ethnicity, site, soil_ts
  - hca_motion_fd_summary.csv    Output of fmri_qc_motion_fd.Rmd; contains RS.fMRI_Count
                                 and TripleOut_FD05_60Fr
  - ssaga_cover_demo01.csv       SSAGA sociodemographics;
                                 household size, marital status, work status, income
  - moca01.csv                   Montreal Cognitive Assessment
  - prsi01.csv                   NIH Toolbox Social Isolation survey

Outputs:
  - datainfo.csv                 Main dataframe: demographics + all recoded variables

Dependencies: tidyverse

Usage:        Update paths in USER CONFIGURATION block; then knit or run chunks sequentially.
-->


# Libraries
```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
```


# USER CONFIGURATION
```{r config}
# Directory containing all raw HCP-A NDA text files
data_dir   <- "/path/to/hca_behavioral_data"
output_dir <- "/path/to/output"

# Input files (all use 2-row NDA header: row 1 = column names, row 2 = descriptions)
demographics_file <- file.path(data_dir, "ndar_subject01.csv")
# hca_motion_fd_summary.csv is produced by fmri_qc_motion_fd.Rmd
motion_qc_file    <- file.path(output_dir, "hca_motion_fd_summary.csv")
ssaga_file        <- file.path(data_dir, "ssaga_cover_demo01.csv")
moca_file         <- file.path(data_dir, "moca01.csv")
prsi_file         <- file.path(data_dir, "prsi01.csv")

# Output file
output_datainfo_file <- file.path(output_dir, "datainfo.csv")
```


# Load demographics and create age variables
```{r demographics}
# NDA files use a 2-row header: row 1 = column names, row 2 = descriptions
# Load column names from row 1, then reload data skipping both header rows
temp     <- read.csv(demographics_file, header = TRUE)
headers  <- colnames(temp)
datainfo <- read.csv(demographics_file, skip = 2, header = FALSE)
colnames(datainfo) <- headers

# Convert all character columns to factor (columns 8 onward: sex, race, ethnicity, site)
for (cc in 8:dim(datainfo)[2]) {
  if (class(datainfo[, cc]) == "character") {
    datainfo[, cc] <- as.factor(datainfo[, cc])
  }
}

# Age in years
datainfo <- datainfo %>%
  mutate(age_yrs = interview_age / 12)

# Age bins: 5 groups combined at 75+ (used in model covariates and group plots)
# Breaks: [36–45), [45–55), [55–65), [65–75), [75+]
datainfo <- datainfo %>%
  mutate(age.bins.3 = cut(
    interview_age / 12,
    breaks = c(
      min(interview_age / 12),
      min(interview_age / 12) + 9,
      min(interview_age / 12) + 19,
      min(interview_age / 12) + 29,
      min(interview_age / 12) + 39,
      max(interview_age / 12)
    ),
    include.lowest = TRUE
  ))
```


# Load scan completeness and motion QC flags
```{r scan_qc}
# hca_motion_fd_summary.csv is produced by fmri_qc_motion_fd.Rmd.
# Motion exclusion criterion: high-motion frames were defined by FD > 0.5 mm, with the
# frame before and after also marked as high-motion (triple-frame exclusion). Participants
# were excluded if any of their four resting-state runs had more than 40% high-motion frames.
# TripleOut_FD05_60Fr == 1 indicates a participant passed this criterion across all four runs.
temp       <- read.csv(motion_qc_file, header = TRUE)
headers    <- colnames(temp)
compl.data <- read.csv(motion_qc_file, skip = 2, header = FALSE)
colnames(compl.data) <- headers

# Drop columns already present in datainfo
drop        <- c("subjectkey", "interview_date", "interview_age", "sex", "visit", "unrelated_subset")
compl.data2 <- compl.data[, !colnames(compl.data) %in% drop]

datainfo <- merge(datainfo, compl.data2, by = "src_subject_id", all.x = TRUE)
```


# Load SSAGA sociodemographics
```{r ssaga_load}
temp       <- read.csv(ssaga_file, header = TRUE)
headers    <- colnames(temp)
ssaga.data <- read.csv(ssaga_file, skip = 2, header = FALSE)
colnames(ssaga.data) <- headers
```

## Household size
```{r household}
# household.2g: 1-person (living alone) vs. 2+person; reference = "2+person" (coded 1)
# household.3g: 1-person, 2-person, 3+person
# 0-person and NA coded as NA
ssaga.data <- ssaga.data %>%
  mutate(
    household.3g = ifelse(
      !is.na(household_number_in_house) & household_number_in_house != 0,
      case_when(
        household_number_in_house == 1 ~ 1,
        household_number_in_house == 2 ~ 2,
        household_number_in_house >= 3 ~ 3
      ),
      NA
    ),
    # 2+person coded as 1 (reference); 1-person coded as 2
    household.2g = ifelse(
      !is.na(household_number_in_house) & household_number_in_house != 0,
      case_when(
        household_number_in_house == 1  ~ 2,
        household_number_in_house >= 2  ~ 1
      ),
      NA
    )
  ) %>%
  mutate(
    household.3g = cut(household.3g, 3, labels = c("1-person", "2-person", "3+person")),
    household.2g = cut(household.2g, 2, labels = c("2+person", "1-person"))
  )

# Remove any pre-existing household.2g (may be in ndar_subject01.csv) and merge custom coding
datainfo <- datainfo %>%
  select(-any_of("household.2g"))

datainfo <- merge(datainfo, ssaga.data[, c("src_subject_id", "household.2g")],
                  by = "src_subject_id", all.x = TRUE)
```

## Marital status
```{r marital}
# marital.2g: Married (includes "Living as Married") vs. Not Married; reference = "Married"
ssaga.data <- ssaga.data %>%
  mutate(across(das1ms, na_if, "")) %>%
  mutate(across(das1ms, as.factor))

ssaga.data$marital.2g <- recode_factor(
  ssaga.data$das1ms,
  "DIVORCED"          = "Not Married",
  "LIVING AS MARRIED" = "Married",
  "MARRIED"           = "Married",
  "NEVER MARRIED"     = "Not Married",
  "SEPARATED"         = "Not Married",
  "WIDOWED"           = "Not Married"
)

datainfo <- merge(datainfo, ssaga.data[, c("src_subject_id", "das1ms", "marital.2g")],
                  by = "src_subject_id", all.x = TRUE)

# Re-code with Married as reference (coded as 1)
datainfo <- datainfo %>%
  mutate(marital.2g = ifelse(
    !is.na(marital.2g),
    case_when(
      marital.2g == "Married"     ~ 1,
      marital.2g == "Not Married" ~ 2
    ),
    NA
  )) %>%
  mutate(marital.2g = cut(marital.2g, 2, labels = c("Married", "Not Married")))
```

## Work status
```{r work}
# work.2g: Employed vs. Unemployed
# Based on eo4 (currently employed) and a2b_yn (working full-time)
# Note: 1 subject reported Unemployed (eo4 = 0) but also full-time (a2b_yn = 1) → coded as NA
ssaga.data <- ssaga.data %>%
  mutate(across(a2b_yn, as.factor)) %>%
  mutate(across(eo4,    as.factor))

ssaga.data <- ssaga.data %>%
  mutate(work.3g = ifelse(
    !is.na(eo4),
    ifelse(!is.na(a2b_yn),
           case_when(
             eo4 == "1" & a2b_yn == "1" ~ 1,
             eo4 == "1" & a2b_yn == "0" ~ 2
           ),
           case_when(eo4 == "0" ~ 3)),
    NA
  )) %>%
  mutate(work.3g = cut(work.3g, 3,
                       labels = c("Employed-Fulltime", "Employed-NotFulltime", "Unemployed")))

ssaga.data <- ssaga.data %>%
  mutate(work.2g = ifelse(
    !is.na(eo4),
    ifelse(eo4 == "0",
           ifelse(is.na(a2b_yn),
                  case_when(TRUE ~ 2),
                  NA),
           case_when(eo4 == "1" ~ 1)),
    NA
  )) %>%
  mutate(work.2g = cut(work.2g, 2, labels = c("Employed", "Unemployed")))

datainfo <- merge(datainfo, ssaga.data[, c("src_subject_id", "eo4", "a2b_yn", "work.3g", "work.2g")],
                  by = "src_subject_id", all.x = TRUE)
```

## Annual family income (log-transformed)
```{r income}
# Log-transform annual family income (natural log)
# 0 and NA → NA (3 subjects reported $0 income)
ssaga.data <- ssaga.data %>%
  mutate(annual_fam_inc.log = ifelse(
    !is.na(annual_fam_inc),
    ifelse(annual_fam_inc != 0, log(annual_fam_inc), NA),
    NA
  ))

# Remove pre-existing income columns (may be in ndar_subject01.csv) and merge updated values
datainfo <- datainfo %>%
  select(-any_of(c("annual_fam_inc", "annual_fam_inc.log")))

datainfo <- merge(datainfo, ssaga.data[, c("src_subject_id", "annual_fam_inc", "annual_fam_inc.log")],
                  by = "src_subject_id", all.x = TRUE)
```


# Load MoCA: cognitive assessment and education groups
```{r moca_load}
temp      <- read.csv(moca_file, header = TRUE)
headers   <- colnames(temp)
moca.data <- read.csv(moca_file, skip = 2, header = FALSE)
colnames(moca.data) <- headers

# Merge MoCA total score into datainfo
datainfo <- merge(datainfo, moca.data[c("src_subject_id", "moca_total")],
                  by = "src_subject_id", all.x = TRUE)
```

```{r education}
# Education groups from MoCA years-of-education variable (moca_edu = actual years)
# Groups: <= 14 yrs → "High School or Below"; 15–18 → "College"; >= 19 → "Graduate"
hca.education <- moca.data %>%
  select(src_subject_id, moca_edu) %>%
  mutate(moca_edu.f = as.factor(moca_edu)) %>%
  mutate(moca_edu.3g = ifelse(
    !is.na(moca_edu.f),
    case_when(
      moca_edu.f %in% c("7", "8", "9", "10", "11", "12", "13", "14") ~ 1,
      moca_edu.f %in% c("15", "16", "17", "18")                       ~ 2,
      TRUE                                                              ~ 3
    ),
    NA
  )) %>%
  mutate(moca_edu.3g = cut(moca_edu.3g, 3,
                           labels = c("High School or Below", "College", "Graduate")))

# Cross-reference: education from SSAGA bkgrnd_education (self-reported categorical)
hca.education <- merge(hca.education, ssaga.data[, c("src_subject_id", "bkgrnd_education")],
                       by = "src_subject_id", all.x = TRUE)
hca.education <- hca.education %>%
  mutate(bkgrnd_education = ifelse(bkgrnd_education == "", NA, bkgrnd_education)) %>%
  mutate(bkgrnd_education.f = as.factor(bkgrnd_education)) %>%
  mutate(bkgrnd_education.3g = ifelse(
    !is.na(bkgrnd_education.f),
    case_when(
      bkgrnd_education.f %in% c("3RD", "9TH", "10TH", "11TH", "12TH")                  ~ 1,
      bkgrnd_education.f %in% c("TECHNICAL SCHOOL OR 1 YR COLLEGE", "2 YRS COLLEGE",
                                  "3 YRS COLLEGE", "4 YRS COLLEGE: B.A., B.S.")          ~ 2,
      TRUE                                                                                ~ 3
    ),
    NA
  )) %>%
  mutate(bkgrnd_education.3g = cut(bkgrnd_education.3g, 3,
                                   labels = c("High School or Below", "College", "Graduate")))

datainfo <- merge(datainfo, hca.education[, c("src_subject_id", "moca_edu.3g", "bkgrnd_education.3g")],
                  by = "src_subject_id", all.x = TRUE)
```


# Load PRSI: NIH Toolbox Social Isolation survey (soil_ts)
```{r prsi_load}
temp       <- read.csv(prsi_file, header = TRUE)
headers    <- colnames(temp)
prsi.data  <- read.csv(prsi_file, skip = 2, header = FALSE)
colnames(prsi.data) <- headers

# Drop metadata columns; retain src_subject_id and survey variables including soil_ts
drop       <- c("collection_id", "prsi01_id", "dataset_id", "subjectkey",
                "interview_date", "interview_age", "sex", "collection_title",
                "interview_language", "version_form")
prsi.data2 <- prsi.data[, !colnames(prsi.data) %in% drop]

datainfo <- merge(datainfo, prsi.data2, by = "src_subject_id", all.x = TRUE)
```


# Loneliness grouping from SOIL T-score
```{r loneliness_groups}
# Loneliness groups based on NIH Toolbox cutoffs: <= 40 = Low; 41–59 = Mid; >= 60 = High
datainfo <- datainfo %>%
  mutate(soil.3g = ifelse(
    !is.na(soil_ts),
    case_when(
      soil_ts >= 60 ~ 3,
      soil_ts >  40 ~ 2,
      soil_ts <= 40 ~ 1
    ),
    NA
  )) %>%
  mutate(soil.3g = cut(soil.3g, 3, labels = c("Lowlonely", "Midlonely", "Highlonely")))
```


# Export main dataframe
```{r export}
write.csv(datainfo, file = output_datainfo_file, row.names = FALSE)
```
