---
title: "Behavioral Analysis: Loneliness and Sociodemographic Factors"
output: html_notebook
---

<!--
Purpose:      Descriptive statistics and regression models examining loneliness
              (SOIL T-score) in relation to sociodemographic factors in the
              final N = 512 sample. Produces manuscript Table 1 & 2 summary stats
              and the Age–Loneliness scatterplot with marginal histograms.

Inputs:
  - datainfo.csv           Output of behav_preproc_recode_variables.Rmd;
                           contains all recoded sociodemographic variables,
                           loneliness, and motion QC flags

Outputs:
  - soil_age_scatter_marginal.tiff    Age–Loneliness scatterplot with marginal histograms

Dependencies: tidyverse, broom, gridExtra, car, ggExtra, rstatix

Usage:        Update paths in USER CONFIGURATION block; then knit or run
              chunks sequentially.
-->


# Libraries
```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(broom)
library(gridExtra)
library(car)
library(ggExtra)
library(rstatix)
```


# USER CONFIGURATION
```{r config}
# Output of behav_preproc_recode_variables.Rmd
datainfo_file <- "/path/to/output/datainfo.csv"

# Directory for saving figures
fig_dir <- "/path/to/figures"

# Output figure file
output_scatter_file <- file.path(fig_dir, "soil_age_scatter_marginal.tiff")
```


# Load data and create analysis sample (N = 512)
```{r load_data}
datainfo <- read.csv(datainfo_file, header = TRUE)

# Final analysis sample (N = 512):
#   - has loneliness score (soil_ts)
#   - passed motion QC (TripleOut_FD05_60Fr == 1)
#   - completed all four resting-state runs (RS.fMRI_Count == 4)
#   - one subject excluded per data quality review
mod9_data <- datainfo %>%
  filter(!is.na(soil_ts),
         TripleOut_FD05_60Fr == 1,
         RS.fMRI_Count == 4,
         src_subject_id != "HCA7083065")

dim(mod9_data)
```


# Mean-center annual family income
```{r income_center}
# Mean-center log-transformed income across the N = 512 sample
# 0-income subjects were coded as NA in behav_preproc_recode_variables.Rmd
mod9_data <- mod9_data %>%
  mutate(annual_fam_inc.log.c = scale(annual_fam_inc.log, center = TRUE, scale = FALSE)) %>%
  mutate(annual_fam_inc.log.c = as.vector(annual_fam_inc.log.c))
```


# Descriptive statistics

## Loneliness
```{r desc_loneliness}
mod9_data %>%
  summarise(n         = n(),
            mean      = mean(soil_ts, na.rm = TRUE),
            sd        = sd(soil_ts, na.rm = TRUE),
            min       = min(soil_ts, na.rm = TRUE),
            max       = max(soil_ts, na.rm = TRUE),
            n_missing = sum(is.na(soil_ts)))
```

## Age
```{r desc_age}
mod9_data %>%
  summarise(n         = n(),
            mean      = mean(age_yrs, na.rm = TRUE),
            sd        = sd(age_yrs, na.rm = TRUE),
            min       = min(age_yrs, na.rm = TRUE),
            max       = max(age_yrs, na.rm = TRUE),
            n_missing = sum(is.na(age_yrs)))
```

## Sex
```{r desc_sex}
str(mod9_data$sex)

mod9_data %>%
  group_by(sex) %>%
  summarise(n = n())

# Sex differences in loneliness
mod9_data %>%
  group_by(sex) %>%
  summarise(n         = n(),
            mean      = mean(soil_ts, na.rm = TRUE),
            sd        = sd(soil_ts, na.rm = TRUE),
            min       = min(soil_ts, na.rm = TRUE),
            max       = max(soil_ts, na.rm = TRUE),
            n_missing = sum(is.na(soil_ts)))

# Variance test: non-significant → equal variance
var.test(data = mod9_data, soil_ts ~ sex)
# t-test (equal variance)
# Note: t.test uses group 2 as reference; lm() uses group 1 (Female) as reference
t.test(data = mod9_data, soil_ts ~ sex, var.equal = TRUE)
summary(lm(data = mod9_data, soil_ts ~ sex))
confint(lm(data = mod9_data, soil_ts ~ sex))
```

## Household
```{r desc_household}
str(mod9_data$household.2g)
levels(mod9_data$household.2g)

mod9_data %>%
  group_by(household.2g) %>%
  summarise(n = n())

# Household differences in loneliness
mod9_data %>%
  group_by(household.2g) %>%
  summarise(n         = n(),
            mean      = mean(soil_ts, na.rm = TRUE),
            sd        = sd(soil_ts, na.rm = TRUE),
            min       = min(soil_ts, na.rm = TRUE),
            max       = max(soil_ts, na.rm = TRUE),
            n_missing = sum(is.na(soil_ts)))

# Variance test: non-significant → equal variance
var.test(data = mod9_data, soil_ts ~ household.2g)
t.test(data = mod9_data, soil_ts ~ household.2g, var.equal = TRUE)
```

## Marital status
```{r desc_marital}
str(mod9_data$marital.2g)
levels(mod9_data$marital.2g)
# Relabel for display
levels(mod9_data$marital.2g) <- c("Married", "Unmarried")

# Marital differences in loneliness
mod9_data %>%
  group_by(marital.2g) %>%
  summarise(n         = n(),
            mean      = mean(soil_ts, na.rm = TRUE),
            sd        = sd(soil_ts, na.rm = TRUE),
            min       = min(soil_ts, na.rm = TRUE),
            max       = max(soil_ts, na.rm = TRUE),
            n_missing = sum(is.na(soil_ts)))

# Variance test: significant → unequal variance; use Welch's t-test
var.test(data = mod9_data, soil_ts ~ marital.2g)
```

## Work status
```{r desc_work}
str(mod9_data$work.2g)

# Work differences in loneliness
mod9_data %>%
  group_by(work.2g) %>%
  summarise(n         = n(),
            mean      = mean(soil_ts, na.rm = TRUE),
            sd        = sd(soil_ts, na.rm = TRUE),
            min       = min(soil_ts, na.rm = TRUE),
            max       = max(soil_ts, na.rm = TRUE),
            n_missing = sum(is.na(soil_ts)))

# Variance test: non-significant → equal variance
var.test(data = mod9_data, soil_ts ~ work.2g)
```

## Education
```{r desc_education}
str(mod9_data$moca_edu.3g)
levels(mod9_data$moca_edu.3g)

# Education differences in loneliness
mod9_data %>%
  group_by(moca_edu.3g) %>%
  summarise(n         = n(),
            mean      = mean(soil_ts, na.rm = TRUE),
            sd        = sd(soil_ts, na.rm = TRUE),
            min       = min(soil_ts, na.rm = TRUE),
            max       = max(soil_ts, na.rm = TRUE),
            n_missing = sum(is.na(soil_ts)))

# Bartlett test for multiple groups: significant → unequal variance
bartlett.test(data = mod9_data, soil_ts ~ moca_edu.3g)
```

## Annual family income
```{r desc_income}
summary(mod9_data$annual_fam_inc)
summary(mod9_data$annual_fam_inc.log)
sum(!is.na(mod9_data$annual_fam_inc.log))

# Exclude 0-income subjects (treated as NA) for summary stats
mod9_data %>%
  filter(annual_fam_inc != 0) %>%
  summarise(mean = mean(annual_fam_inc), sd = sd(annual_fam_inc))
```


# Regression models
```{r models}
# Pearson r: age–loneliness
mod9_data %>%
  summarise(r       = cor.test(soil_ts, age_yrs)[["estimate"]],
            p.value = cor.test(soil_ts, age_yrs)[["p.value"]])

mod9.models <- mod9_data %>%
  nest() %>%
  mutate(
    # Model 0: Age
    model0     = map(data, ~ lm(soil_ts ~ age_yrs, data = .)),
    # Pearson r (standardized)
    model0.1   = map(data, ~ lm(scale(soil_ts) ~ scale(age_yrs), data = .)),
    # Model 1: Sex (equal variance)
    # lm() takes Group 1 (Female) as reference
    model1     = map(data, ~ lm(soil_ts ~ sex, data = .)),
    model1.1   = map(data, ~ Anova(lm(soil_ts ~ sex, data = .), type = "III")),
    # Sex × Age
    model1.2   = map(data, ~ lm(soil_ts ~ age_yrs * sex, data = .)),
    # Model 2: Household (equal variance)
    model2     = map(data, ~ lm(soil_ts ~ household.2g, data = .)),
    model2.1   = map(data, ~ Anova(lm(soil_ts ~ household.2g, data = .), type = "III")),
    # Household × Age
    model2.2   = map(data, ~ lm(soil_ts ~ age_yrs * household.2g, data = .)),
    # Household × Sex
    model2.3   = map(data, ~ lm(soil_ts ~ sex * household.2g, data = .)),
    model2.3.1 = map(data, ~ Anova(lm(soil_ts ~ sex * household.2g, data = .), type = "III")),
    # Model 3: Marital (unequal variance — use oneway.test / Welch's ANOVA)
    # Note: t.test CIs have opposite reference group to lm()
    model3     = map(data, ~ lm(soil_ts ~ marital.2g, data = .)),
    model3.1   = map(data, ~ oneway.test(soil_ts ~ marital.2g, data = .)),
    # Marital × Age
    model3.2   = map(data, ~ lm(soil_ts ~ age_yrs * marital.2g, data = .)),
    # Marital × Sex
    model3.3   = map(data, ~ lm(soil_ts ~ sex * marital.2g, data = .)),
    model3.3.1 = map(data, ~ Anova(lm(soil_ts ~ sex * marital.2g, data = .), type = "III")),
    # Model 4: Full sociodemographic model
    model4     = map(data, ~ lm(soil_ts ~ age_yrs + sex + household.2g + marital.2g +
                                           work.2g + annual_fam_inc.log.c + moca_edu.3g, data = .)),
    # Mean-centered age
    model4.1   = map(data, ~ lm(soil_ts ~ scale(age_yrs, center = TRUE, scale = FALSE) +
                                           sex + household.2g + marital.2g +
                                           work.2g + annual_fam_inc.log.c + moca_edu.3g, data = .)),
    # Standardized age and income — manuscript Table 2
    model4.2   = map(data, ~ lm(scale(soil_ts) ~ scale(age_yrs, center = TRUE, scale = TRUE) +
                                                   sex + household.2g + marital.2g +
                                                   work.2g + scale(annual_fam_inc.log, center = TRUE, scale = TRUE) +
                                                   moca_edu.3g, data = .)),
    # Model 5: Three-way Age × Sex × Marital
    model5     = map(data, ~ lm(soil_ts ~ age_yrs * marital.2g * sex, data = .)),
    model5.1   = map(data, ~ Anova(lm(soil_ts ~ age_yrs * marital.2g * sex, data = .), type = "III")),
    # Three-way + remaining covariates
    model5.2   = map(data, ~ Anova(lm(soil_ts ~ age_yrs * marital.2g * sex +
                                                 household.2g + work.2g + annual_fam_inc.log.c + moca_edu.3g, data = .),
                                   type = "III")),
    # Model 6: Work (equal variance)
    model6     = map(data, ~ lm(soil_ts ~ work.2g, data = .)),
    model6.1   = map(data, ~ Anova(lm(soil_ts ~ work.2g, data = .), type = "III")),
    # Work × Age
    model6.2   = map(data, ~ lm(soil_ts ~ age_yrs * work.2g, data = .)),
    # Work × Sex
    model6.3   = map(data, ~ lm(soil_ts ~ sex * work.2g, data = .)),
    model6.3.1 = map(data, ~ Anova(lm(soil_ts ~ sex * work.2g, data = .), type = "III")),
    # Model 7: Income
    model7     = map(data, ~ lm(soil_ts ~ annual_fam_inc.log.c, data = .)),
    # Standardized coefficient (= Pearson r)
    model7.0   = map(data, ~ lm(scale(soil_ts) ~ scale(annual_fam_inc.log), data = .)),
    # Income × Age
    model7.1   = map(data, ~ lm(soil_ts ~ age_yrs * annual_fam_inc.log.c, data = .)),
    # Income × Sex
    model7.2   = map(data, ~ lm(soil_ts ~ sex * annual_fam_inc.log.c, data = .)),
    # Model 8: Education (unequal variance — use oneway.test / white.adjust)
    model8     = map(data, ~ lm(soil_ts ~ moca_edu.3g, data = .)),
    model8.1   = map(data, ~ Anova(lm(soil_ts ~ moca_edu.3g, data = .), type = "III",
                                   white.adjust = TRUE)),
    model8.1.1 = map(data, ~ oneway.test(soil_ts ~ moca_edu.3g, data = .)),
    # Education × Age
    model8.2   = map(data, ~ lm(soil_ts ~ age_yrs * moca_edu.3g, data = .)),
    # Education × Sex
    model8.3   = map(data, ~ lm(soil_ts ~ sex * moca_edu.3g, data = .)),
    model8.3.1 = map(data, ~ Anova(lm(soil_ts ~ sex * moca_edu.3g, data = .), type = "III")),
    # broom::tidy — regression coefficients with 95% CIs
    tidied0     = map(model0,     tidy, conf.int = TRUE, conf.level = 0.95),
    tidied0.1   = map(model0.1,   tidy, conf.int = TRUE, conf.level = 0.95),
    glance0     = map(model0,     glance),
    tidied1     = map(model1,     tidy, conf.int = TRUE, conf.level = 0.95),
    glance1     = map(model1,     glance),
    tidied1.1   = map(model1.1,   tidy, conf.int = TRUE, conf.level = 0.95),
    tidied1.2   = map(model1.2,   tidy, conf.int = TRUE, conf.level = 0.95),
    tidied2     = map(model2,     tidy, conf.int = TRUE, conf.level = 0.95),
    glance2     = map(model2,     glance),
    tidied2.1   = map(model2.1,   tidy, conf.int = TRUE, conf.level = 0.95),
    tidied2.2   = map(model2.2,   tidy, conf.int = TRUE, conf.level = 0.95),
    tidied2.3   = map(model2.3,   tidy, conf.int = TRUE, conf.level = 0.95),
    tidied2.3.1 = map(model2.3.1, tidy, conf.int = TRUE, conf.level = 0.95),
    tidied3     = map(model3,     tidy, conf.int = TRUE, conf.level = 0.95),
    glance3     = map(model3,     glance),
    tidied3.1   = map(model3.1,   tidy, conf.int = TRUE, conf.level = 0.95),
    tidied3.2   = map(model3.2,   tidy, conf.int = TRUE, conf.level = 0.95),
    tidied3.3   = map(model3.3,   tidy, conf.int = TRUE, conf.level = 0.95),
    tidied3.3.1 = map(model3.3.1, tidy, conf.int = TRUE, conf.level = 0.95),
    tidied4     = map(model4,     tidy, conf.int = TRUE, conf.level = 0.95),
    glance4     = map(model4,     glance),
    tidied4.1   = map(model4.1,   tidy, conf.int = TRUE, conf.level = 0.95),
    glance4.1   = map(model4.1,   glance),
    tidied4.2   = map(model4.2,   tidy, conf.int = TRUE, conf.level = 0.95),
    glance4.2   = map(model4.2,   glance),
    tidied5     = map(model5,     tidy, conf.int = TRUE, conf.level = 0.95),
    glance5     = map(model5,     glance),
    tidied5.1   = map(model5.1,   tidy, conf.int = TRUE, conf.level = 0.95),
    tidied5.2   = map(model5.2,   tidy, conf.int = TRUE, conf.level = 0.95),
    glance6     = map(model6,     glance),
    tidied6     = map(model6,     tidy, conf.int = TRUE, conf.level = 0.95),
    tidied6.1   = map(model6.1,   tidy, conf.int = TRUE, conf.level = 0.95),
    tidied6.2   = map(model6.2,   tidy, conf.int = TRUE, conf.level = 0.95),
    tidied6.3   = map(model6.3,   tidy, conf.int = TRUE, conf.level = 0.95),
    tidied6.3.1 = map(model6.3.1, tidy, conf.int = TRUE, conf.level = 0.95),
    glance7     = map(model7,     glance),
    tidied7     = map(model7,     tidy, conf.int = TRUE, conf.level = 0.95),
    tidied7.0   = map(model7.0,   tidy, conf.int = TRUE, conf.level = 0.95),
    tidied7.1   = map(model7.1,   tidy, conf.int = TRUE, conf.level = 0.95),
    tidied7.2   = map(model7.2,   tidy, conf.int = TRUE, conf.level = 0.95),
    glance8     = map(model8,     glance),
    tidied8     = map(model8,     tidy, conf.int = TRUE, conf.level = 0.95),
    tidied8.1   = map(model8.1,   tidy, conf.int = TRUE, conf.level = 0.95),
    tidied8.1.1 = map(model8.1.1, tidy, conf.int = TRUE, conf.level = 0.95),
    tidied8.2   = map(model8.2,   tidy, conf.int = TRUE, conf.level = 0.95),
    tidied8.3   = map(model8.3,   tidy, conf.int = TRUE, conf.level = 0.95),
    tidied8.3.1 = map(model8.3.1, tidy, conf.int = TRUE, conf.level = 0.95)
  ) %>%
  select(-data)
```


# Post-hoc tests

## Marital: Welch's t-test (unequal variance)
```{r posthoc_marital}
# Note: t.test reference group is OPPOSITE to lm() — group 2 (Unmarried) is reference
t.test(data = mod9_data, soil_ts ~ marital.2g, var.equal = FALSE)
```

## Education: Games-Howell post-hoc test (unequal variance, 3 groups)
```{r posthoc_education}
games_howell_test(data = mod9_data, soil_ts ~ moca_edu.3g, conf.level = 0.95)
```


# Age–Loneliness scatterplot with marginal histograms
```{r scatter_marginal}
# Loess fit (panel a)
p.loess <- mod9_data %>%
  ggplot(aes(x = age_yrs, y = soil_ts)) +
  geom_point() +
  geom_smooth(method = "loess") +
  labs(x = "Age (years)", y = "Loneliness") +
  ggtitle("(a)")

p.loess.marginal <- ggMarginal(p.loess,
                               type    = "histogram",
                               xparams = list(fill = "white", col = "skyblue4"),
                               yparams = list(fill = "white", col = "springgreen4"))

# Linear fit (panel b)
p.lm <- mod9_data %>%
  ggplot(aes(x = age_yrs, y = soil_ts)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(x = "Age (years)") +
  theme(axis.title.y = element_blank()) +
  ggtitle("(b)")

p.lm.marginal <- ggMarginal(p.lm,
                             type    = "histogram",
                             xparams = list(fill = "white", col = "skyblue4"),
                             yparams = list(fill = "white", col = "springgreen4"))

# Combine and save
p.marginal.combine <- arrangeGrob(p.loess.marginal, p.lm.marginal, ncol = 2)
ggsave(output_scatter_file, p.marginal.combine, dpi = 300)
```
