---
title: "HCP-Aging QC: Framewise Displacement"
author: "Samantha Chen"
date: "Feb 02 2022"
output: html_notebook
---

## Overview

This notebook computes and visualizes framewise displacement (FD) for quality
control of HCP-Aging (HCA) resting-state and task fMRI data.

For each subject and functional run, the notebook:

1. Reads `Movement_Regressors.txt` from the HCP pipeline output
2. Plots 6 motion parameters (3 translations + 3 rotations)
3. Calculates FD using the Power et al. (2012) formula
4. Flags frames exceeding FD thresholds (0.3 mm and 0.5 mm)
5. Saves per-run FD timeseries and combined motion + FD plots as PDF

A group-level FD summary CSV is saved at the end.

**Inputs:**

- HCA fMRI results directory (`hca_mmp`): contains `Movement_Regressors.txt` per run
- Subject list CSV: with column `src_subject_id` (format: `HCA<numeric_id>`)

**Outputs:**

- Per-run FD timeseries: `<output_motion>/HCA<subj>/HCA<subj>_<run>_framewise.csv`
- Per-run motion + FD plot: `<output_motion>/HCA<subj>/HCA<subj>_<run>_motion.pdf`
- Group FD summary: `<output_motion>/Framewise_summary.csv`


```{r setup, message=FALSE}
library(tidyverse)  # dplyr, ggplot2
library(reshape)    # melt() for wide-to-long conversion
library(gridExtra)  # combine multiple ggplot panels
```


```{r configuration}
# ── USER CONFIGURATION ─────────────────────────────────────────────────────────
# Update these paths before running.

# Root directory of HCA fmriresults01 data
hca_mmp <- "/path/to/HCA_data/fmriresults01"

# Output directory for motion QC files and plots
output_motion <- "/path/to/output/motion_qc"

# Subject list CSV (must contain column: src_subject_id, format HCA<numeric_id>)
subj_list_file <- "/path/to/HCA_subject_list.csv"

# HCA functional run names (as named in the HCP directory structure)
hca_func_list <- list(
  "rfMRI_REST1_AP", "rfMRI_REST1_PA",
  "rfMRI_REST2_AP", "rfMRI_REST2_PA",
  "tfMRI_CARIT_PA", "tfMRI_FACENAME_PA", "tfMRI_VISMOTOR_PA"
)

# Column names for the 12-column Movement_Regressors.txt (6 params + 6 derivatives)
# HCP pipeline format: X Y Z RotX RotY RotZ dX dY dZ dRotX dRotY dRotZ
motion_colnames <- c("X", "Y", "Z", "RotX", "RotY", "RotZ",
                     "dX", "dY", "dZ", "dRotX", "dRotY", "dRotZ")

# FD outlier thresholds (mm)
fd_thr_low  <- 0.3
fd_thr_high <- 0.5
# ───────────────────────────────────────────────────────────────────────────────
```


```{r load_subjects}
# Read subject list and extract the numeric part of each subject ID
# src_subject_id format: "HCA<numeric_id>"
subj_info <- read.csv(subj_list_file, header = TRUE)
subjects <- substr(subj_info$src_subject_id, 4, nchar(subj_info$src_subject_id))
```


```{r compute_fd}
# Initialize group-level FD summary data frame
FD_dat <- data.frame()

for (ss in 1:length(subjects)) {

  # Create subject output directory if it does not exist
  subj_output <- file.path(output_motion, paste0("HCA", subjects[ss]))
  if (!file.exists(subj_output)) {
    dir.create(subj_output, recursive = TRUE)
  }

  # Locate this subject's directory in the HCA data (matched by numeric ID)
  hca_subject <- grep(list.files(hca_mmp), pattern = subjects[ss], value = TRUE)

  for (tt in 1:length(hca_func_list)) {
    hca_func     <- hca_func_list[[tt]]
    hca_func_dir <- file.path(hca_mmp, hca_subject, "MNINonLinear", "Results", hca_func)

    # Some subjects did not complete all scans; skip missing runs
    if (!file.exists(hca_func_dir)) next

    # ── Read motion regressors ────────────────────────────────────────────────
    # HCP output: Movement_Regressors.txt, dim [n_frames x 12], no header
    motion_hca_temp <- read.table(
      file.path(hca_func_dir, "Movement_Regressors.txt"),
      header = FALSE, sep = ""
    )
    colnames(motion_hca_temp) <- motion_colnames

    # ── Plot 6 motion parameters ──────────────────────────────────────────────
    # Reshape to long format for ggplot
    temp_hca_df <- data.frame(x = seq_along(motion_hca_temp[, 1]),
                              motion_hca_temp[, 1:6])
    temp_hca_df_long <- melt(temp_hca_df, id.vars = "x")

    hca_motion_reg <- ggplot(temp_hca_df_long, aes(x = x, y = value, color = variable)) +
      geom_line() +
      ggtitle(paste("HCA sub", subjects[ss], hca_func)) +
      labs(x = "Frame #", y = "Trans. (mm) & Rot. (degrees)") +
      coord_cartesian(ylim = c(-2, 2)) +
      theme(legend.title    = element_blank(),
            legend.position = "top",
            legend.box      = "horizontal") +
      guides(color = guide_legend(nrow = 1))

    # ── Calculate framewise displacement (Power et al., 2012) ─────────────────
    # FD = sum of absolute frame-to-frame displacements across 6 motion parameters.
    # Rotations (degrees) are converted to mm assuming a 50 mm sphere radius.
    temp_hca_fd <- motion_hca_temp

    hca_trans  <- cbind(c(0, diff(temp_hca_fd[, 1])),
                        c(0, diff(temp_hca_fd[, 2])),
                        c(0, diff(temp_hca_fd[, 3])))
    hca_rot    <- cbind(c(0, diff(temp_hca_fd[, 4])),
                        c(0, diff(temp_hca_fd[, 5])),
                        c(0, diff(temp_hca_fd[, 6])))
    hca_rot_mm <- (pi / 180) * hca_rot * 50

    temp_hca_fd$FD <- apply(abs(cbind(hca_trans, hca_rot_mm)), 1, sum)

    # Build per-run FD data frame (Frame index + FD value)
    hca_fd <- data.frame(Frame = seq_along(temp_hca_fd[, 1]),
                         FD    = temp_hca_fd$FD)

    # Flag outlier frames at each threshold (1 = outlier, 0 = clean)
    hca_fd$FD_outlier_thr03 <- as.integer(hca_fd$FD >= fd_thr_low)
    hca_fd$FD_outlier_thr05 <- as.integer(hca_fd$FD >= fd_thr_high)

    # Save per-run FD timeseries
    write.csv(
      hca_fd,
      file      = file.path(subj_output,
                             paste0("HCA", subjects[ss], "_", hca_func, "_framewise.csv")),
      row.names = FALSE
    )

    # Append per-run summary statistics to group data frame
    temp_fd_dat <- data.frame(
      Subject            = subjects[ss],
      Task               = hca_func,
      FD_average         = mean(hca_fd$FD),
      FD_max             = max(hca_fd$FD),
      FD_SD              = sd(hca_fd$FD),
      FD_noutlier_thr03  = sum(hca_fd$FD_outlier_thr03),
      FD_noutlier_thr05  = sum(hca_fd$FD_outlier_thr05)
    )
    FD_dat <- rbind(FD_dat, temp_fd_dat)

    # ── Plot FD timeseries ────────────────────────────────────────────────────
    hca_fd_plot <- ggplot(data = hca_fd, aes(x = Frame, y = FD)) +
      geom_line() +
      ggtitle(paste("HCA sub", subjects[ss], hca_func)) +
      labs(y = "FD (mm)", x = "Frame #") +
      coord_cartesian(ylim = c(0, 2)) +
      geom_hline(yintercept = c(fd_thr_low, fd_thr_high),
                 color      = c("#00BFC4", "#F8766D"),
                 linetype   = "dashed",
                 linewidth  = 1) +   # cyan = 0.3 mm; red = 0.5 mm
      theme(legend.title    = element_blank(),
            legend.position = "top",
            legend.box      = "horizontal") +
      guides(color = guide_legend(nrow = 1))

    # Save combined motion parameters + FD plot as PDF
    pdf(file.path(subj_output,
                  paste0("HCA", subjects[ss], "_", hca_func, "_motion.pdf")))
    grid.arrange(hca_motion_reg, hca_fd_plot, ncol = 1)
    dev.off()
  }
}

# Save group-level FD summary
write.csv(FD_dat,
          file      = file.path(output_motion, "Framewise_summary.csv"),
          row.names = FALSE)
```
