---
title: "fMRI Preprocessing: Compile Nodal Stats and Prepare PALM Inputs"
output: html_notebook
---

<!--
Purpose:      Compile per-subject nodal graph metrics (absolute-edge variant) into
              merged CSV files, then prepare nodal data files and design matrices
              for PALM permutation testing (Model 9: N = 512; Model 9.4: N = 390).

Inputs:
  - datainfo.csv                   Output of behav_preproc_recode_variables.Rmd;
                                   contains demographic, behavioral, and motion
                                   QC variables (soil_ts, TripleOut_FD05_60Fr,
                                   RS.fMRI_Count, rfMRIxxx_FD_average, moca_total,
                                   sex, site, household.2g, marital.2g, age_yrs)
  - subject_list_file.csv          Subject completion list
  - abs_dir/                       Network graph output directory
  - tgmv_file.csv                  Total gray matter volume
  - net_summary_abs_file           Merged network summary stat CSV; must contain
                                   src_subject_id, modularity_CAnet,
                                   average_shortest_path_length

Outputs — Model 9 (N = 512), written to abs_palm_dir/:
  Nodal PALM inputs (no header, no ID):
    abs_nodal_{metric}_n512.csv
  Global PALM inputs (no header, no ID):
    abs_global_modularity_n512.csv
    abs_global_average_shortest_path_n512.csv
  Design matrix:       abs_design_matrix_n512.csv
  Site block:          abs_site_block_n512.csv
  Verification files:  verify_abs_nodal_{metric}_n512.csv  (with subject ID)

Outputs — Model 9.4 (N = 390), written to abs_palm94_dir/:
  Same naming pattern with n390 suffix.

Dependencies: tidyverse

Usage:        Update all paths in USER CONFIGURATION block. Run the "Compile
              nodal stats" chunk before the PALM input chunks, or ensure the
              in-memory DataFrames (df_strength_abs, etc.) are available.

Notes:
  - Model 9 design matrix columns (12): soil_ts, age_yrs, soilxage, sex_num,
    totalGray, rfMRI_FD_average, moca_total, site_MGH, site_UCLA, site_UMinn,
    site_WashU, intercept
  - Model 9.4 design matrix columns (14): soil_ts, age_yrs, soilxage,
    sex.numeric, household.numeric, marital.numeric, totalGray,
    rfMRI_FD_average, moca_total, site_MGH, site_UCLA, site_UMinn,
    site_WashU, intercept
  - Sex: Female = -1, Male = 1
  - Household: 1-person = 1, 2+person = -1
  - Marital: Not Married = 1, Married = -1
  - Site blocks: MGH = 1, UCLA = 2, UMinn = 3, WashU = 4
-->


# Libraries
```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
```


# USER CONFIGURATION
```{r config}
# ---- Input files ----
# Output of behav_preproc_recode_variables.Rmd
datainfo_file        <- "/path/to/output/datainfo.csv"

# Subject completion list; must have src_subject_id and Pos_only_graph
subject_list_file    <- "/path/to/subject_list.csv"

# Total gray matter volume CSV; must have src_subject_id and totalGray
tgmv_file            <- "/path/to/TotalGray.csv"

# Merged network summary stat CSV (absolute-edge variant)
net_summary_abs_file <- "/path/to/merge_abs_network_summary_stat.csv"

# ---- Network graph directory ----
# Must contain per-subject subdirectories with {subject}_nodal_stats.csv
abs_dir <- "/path/to/network_graph_absolute"

# ---- PALM working directories ----
# Model 9 (N = 512): must match palm_dir in run_fmri_stats_palm.slurm
abs_palm_dir   <- "/path/to/palm_models/abs_model9"

# Model 9.4 (N = 390)
abs_palm94_dir <- "/path/to/palm_models/abs_model9_4"
```


# Load data
```{r load_data}
datainfo <- read.csv(datainfo_file, header = TRUE)

# Subject completion list; filter to subjects with positive-only graph analysis
temp_graph <- read.csv(subject_list_file, header = TRUE)
# Fix BOM encoding artifact in first column name if present
colnames(temp_graph)[1] <- "src_subject_id"

graphdata <- temp_graph %>%
  filter(Pos_only_graph == 1)

# Merge behavioral and demographic variables from datainfo
graphdata <- merge(
  graphdata,
  datainfo[, c("src_subject_id", "age_yrs", "age.bins.3", "sex",
               "household.2g", "marital.2g", "work.2g", "annual_fam_inc.log",
               "moca_edu.3g", "soil_ts", "soil.3g",
               "TripleOut_FD05_60Fr", "RS.fMRI_Count",
               "rfMRI1AP_FD_average", "rfMRI1PA_FD_average",
               "rfMRI2AP_FD_average", "rfMRI2PA_FD_average",
               "moca_total", "site")],
  by = "src_subject_id",
  all.x = TRUE
)

# Base sample: has loneliness score, passed motion QC, completed all 4 RS-fMRI runs
finaldata <- graphdata %>%
  filter(!is.na(soil_ts), TripleOut_FD05_60Fr == 1, RS.fMRI_Count == 4)

dim(finaldata)
```


# Compile nodal stats — Abs
```{r compile_nodal_abs}
# Compile per-subject nodal stats into merged DataFrames.
# Written to abs_dir/ for reference (with subject ID); PALM-ready files are
# written later in the Model 9 / Model 9.4 sections.

df_strength_abs     <- data.frame()
df_normstrength_abs <- data.frame()
df_closecent_abs    <- data.frame()
df_betweencent_abs  <- data.frame()
df_eigencent_abs    <- data.frame()
df_pgcent_abs       <- data.frame()
df_clustercoef_abs  <- data.frame()
df_partcoef_abs     <- data.frame()

for (ii in 1:nrow(finaldata)) {
  subject <- finaldata$src_subject_id[ii]
  nodal_f <- file.path(abs_dir, subject, paste0(subject, "_nodal_stats.csv"))
  read_f  <- read.csv(nodal_f, header = TRUE)

  df_strength_abs[ii, 1]     <- subject
  df_normstrength_abs[ii, 1] <- subject
  df_closecent_abs[ii, 1]    <- subject
  df_betweencent_abs[ii, 1]  <- subject
  df_eigencent_abs[ii, 1]    <- subject
  df_pgcent_abs[ii, 1]       <- subject
  df_clustercoef_abs[ii, 1]  <- subject
  df_partcoef_abs[ii, 1]     <- subject

  colnames(df_strength_abs)[1]     <- "src_subject_id"
  colnames(df_normstrength_abs)[1] <- "src_subject_id"
  colnames(df_closecent_abs)[1]    <- "src_subject_id"
  colnames(df_betweencent_abs)[1]  <- "src_subject_id"
  colnames(df_eigencent_abs)[1]    <- "src_subject_id"
  colnames(df_pgcent_abs)[1]       <- "src_subject_id"
  colnames(df_clustercoef_abs)[1]  <- "src_subject_id"
  colnames(df_partcoef_abs)[1]     <- "src_subject_id"

  for (cc in seq_along(read_f$strenth))                 { df_strength_abs[ii,     cc + 1] <- read_f$strenth[cc] }
  for (cc in seq_along(read_f$normstrength))            { df_normstrength_abs[ii, cc + 1] <- read_f$normstrength[cc] }
  for (cc in seq_along(read_f$closecent))               { df_closecent_abs[ii,    cc + 1] <- read_f$closecent[cc] }
  for (cc in seq_along(read_f$betweencent))             { df_betweencent_abs[ii,  cc + 1] <- read_f$betweencent[cc] }
  for (cc in seq_along(read_f$eigencent))               { df_eigencent_abs[ii,    cc + 1] <- read_f$eigencent[cc] }
  for (cc in seq_along(read_f$pgcent))                  { df_pgcent_abs[ii,       cc + 1] <- read_f$pgcent[cc] }
  for (cc in seq_along(read_f$clustercoef))             { df_clustercoef_abs[ii,  cc + 1] <- read_f$clustercoef[cc] }
  for (cc in seq_along(read_f$participationcoef_CAnet)) { df_partcoef_abs[ii,     cc + 1] <- read_f$participationcoef_CAnet[cc] }
}

dim(df_normstrength_abs)
dim(df_partcoef_abs)
```


# Load network summary stats
```{r load_net_summary}
df_net_abs <- read.csv(net_summary_abs_file, header = TRUE)
```


# Prepare confound variables
```{r confounds}
# Total gray matter volume
tgmv_data <- read.csv(tgmv_file, header = TRUE)
finaldata  <- merge(finaldata, tgmv_data, by = "src_subject_id", all.x = TRUE)
summary(finaldata$totalGray)

# Average mean FD across 4 runs
finaldata <- finaldata %>%
  rowwise() %>%
  mutate(rfMRI_FD_average = mean(
    c(rfMRI1AP_FD_average, rfMRI1PA_FD_average,
      rfMRI2AP_FD_average, rfMRI2PA_FD_average),
    na.rm = TRUE
  )) %>%
  ungroup()

# Sex: numeric coding Female = -1, Male = 1
finaldata <- finaldata %>%
  mutate(sex_num = case_when(sex == "F" ~ -1,
                             sex == "M" ~  1))

# Household: 1-person = 1, 2+person = -1
finaldata <- finaldata %>%
  mutate(household.numeric = case_when(household.2g == "1-person" ~  1,
                                       household.2g == "2+person" ~ -1))

# Marital: Not Married = 1, Married = -1
finaldata <- finaldata %>%
  mutate(marital.numeric = case_when(marital.2g == "Not Married" ~  1,
                                     marital.2g == "Married"     ~ -1))

# Site dummy coding
table(finaldata$site)
finaldata <- finaldata %>%
  mutate(site_MGH   = ifelse(site == "MGH",   1, 0),
         site_UCLA  = ifelse(site == "UCLA",  1, 0),
         site_UMinn = ifelse(site == "UMinn", 1, 0),
         site_WashU = ifelse(site == "WashU", 1, 0))

# Site block for PALM exchangeability: MGH=1, UCLA=2, UMinn=3, WashU=4
finaldata <- finaldata %>%
  mutate(site_block = case_when(site == "MGH"   ~ 1,
                                site == "UCLA"  ~ 2,
                                site == "UMinn" ~ 3,
                                site == "WashU" ~ 4))

# Intercept column
finaldata$intercept <- 1
```


# Model 9 (N = 512)

## Define sample
```{r mod9_sample}
# N = 512: finaldata excluding one subject removed per data quality review
mod9_data <- finaldata %>%
  filter(src_subject_id != "HCA7083065")

# Re-center soil_ts and age_yrs within the N = 512 sample for interaction term
mod9_data$age_yrs.c <- scale(mod9_data$age_yrs, center = TRUE, scale = FALSE)
mod9_data$soil_ts.c <- scale(mod9_data$soil_ts, center = TRUE, scale = FALSE)
mod9_data <- mod9_data %>%
  mutate(soilxage = soil_ts.c * age_yrs.c)

mod9_subjects <- mod9_data %>% select(src_subject_id)
n9 <- nrow(mod9_data)   # 512
dim(mod9_data)
```

## Nodal PALM files
```{r mod9_nodal}
# Output names match fmri_stats_palm.m expectations:
#   abs_nodal_{metric}_n{N}.csv  (no header, no ID)
# Verification files (with subject ID) written with verify_ prefix.

nodal_dfs_abs <- list(
  strength     = df_strength_abs,
  normstrength = df_normstrength_abs,
  closecent    = df_closecent_abs,
  betweencent  = df_betweencent_abs,
  eigencent    = df_eigencent_abs,
  pgcent       = df_pgcent_abs,
  clustercoef  = df_clustercoef_abs,
  partcoef     = df_partcoef_abs
)

for (metric in names(nodal_dfs_abs)) {
  temp_sub <- merge(mod9_subjects, nodal_dfs_abs[[metric]],
                    by = "src_subject_id", all.x = TRUE)
  out_name <- paste0("abs_nodal_", metric, "_n", n9, ".csv")

  # PALM input: no header, no row names, no ID column
  write.table(temp_sub[, 2:ncol(temp_sub)],
              file.path(abs_palm_dir, out_name),
              row.names = FALSE, col.names = FALSE, sep = ",")
  # Verification file with subject ID
  write.csv(temp_sub,
            file.path(abs_palm_dir, paste0("verify_", out_name)),
            row.names = FALSE)
}
```

## Global PALM files (modularity and average shortest path)
```{r mod9_global}
df_mod <- merge(mod9_subjects,
                df_net_abs[, c("src_subject_id", "modularity_CAnet")],
                by = "src_subject_id", all.x = TRUE)
df_mod %>%
  select(modularity_CAnet) %>%
  write.table(., file.path(abs_palm_dir, paste0("abs_global_modularity_n", n9, ".csv")),
              row.names = FALSE, col.names = FALSE, sep = ",")

df_path <- merge(mod9_subjects,
                 df_net_abs[, c("src_subject_id", "average_shortest_path_length")],
                 by = "src_subject_id", all.x = TRUE)
df_path %>%
  select(average_shortest_path_length) %>%
  write.table(., file.path(abs_palm_dir, paste0("abs_global_average_shortest_path_n", n9, ".csv")),
              row.names = FALSE, col.names = FALSE, sep = ",")
```

## Design matrix and site block
```{r mod9_design}
# Design matrix columns (12):
# soil_ts, age_yrs, soilxage, sex_num, totalGray, rfMRI_FD_average,
# moca_total, site_MGH, site_UCLA, site_UMinn, site_WashU, intercept
mod9_data %>%
  select(soil_ts, age_yrs, soilxage, sex_num, totalGray,
         rfMRI_FD_average, moca_total,
         site_MGH, site_UCLA, site_UMinn, site_WashU,
         intercept) %>%
  write.table(., file.path(abs_palm_dir, paste0("abs_design_matrix_n", n9, ".csv")),
              row.names = FALSE, col.names = FALSE, sep = ",")

mod9_data %>%
  select(site_block) %>%
  write.table(., file.path(abs_palm_dir, paste0("abs_site_block_n", n9, ".csv")),
              row.names = FALSE, col.names = FALSE, sep = ",")
```


# Model 9.4 (N = 390)

## Define sample
```{r mod94_sample}
# N = 390: adds non-missing sex, household.2g, and marital.2g filters
mod94_data <- finaldata %>%
  filter(!is.na(sex), !is.na(household.2g), !is.na(marital.2g),
         src_subject_id != "HCA7083065")

# Re-center soil_ts and age_yrs within the N = 390 sample for interaction term
mod94_data$age_yrs.c <- scale(mod94_data$age_yrs, center = TRUE, scale = FALSE)
mod94_data$soil_ts.c <- scale(mod94_data$soil_ts, center = TRUE, scale = FALSE)
mod94_data <- mod94_data %>%
  mutate(soilxage = soil_ts.c * age_yrs.c)

mod94_subjects <- mod94_data %>% select(src_subject_id)
n94 <- nrow(mod94_data)   # 390
dim(mod94_data)
```

## Nodal PALM files
```{r mod94_nodal}
for (metric in names(nodal_dfs_abs)) {
  temp_sub <- merge(mod94_subjects, nodal_dfs_abs[[metric]],
                    by = "src_subject_id", all.x = TRUE)
  out_name <- paste0("abs_nodal_", metric, "_n", n94, ".csv")

  write.table(temp_sub[, 2:ncol(temp_sub)],
              file.path(abs_palm94_dir, out_name),
              row.names = FALSE, col.names = FALSE, sep = ",")
}
```

## Global PALM files (modularity and average shortest path)
```{r mod94_global}
df_mod <- merge(mod94_subjects,
                df_net_abs[, c("src_subject_id", "modularity_CAnet")],
                by = "src_subject_id", all.x = TRUE)
df_mod %>%
  select(modularity_CAnet) %>%
  write.table(., file.path(abs_palm94_dir, paste0("abs_global_modularity_n", n94, ".csv")),
              row.names = FALSE, col.names = FALSE, sep = ",")

df_path <- merge(mod94_subjects,
                 df_net_abs[, c("src_subject_id", "average_shortest_path_length")],
                 by = "src_subject_id", all.x = TRUE)
df_path %>%
  select(average_shortest_path_length) %>%
  write.table(., file.path(abs_palm94_dir, paste0("abs_global_average_shortest_path_n", n94, ".csv")),
              row.names = FALSE, col.names = FALSE, sep = ",")
```

## Design matrix and site block
```{r mod94_design}
# Design matrix columns (14):
# soil_ts, age_yrs, soilxage, sex.numeric, household.numeric, marital.numeric,
# totalGray, rfMRI_FD_average, moca_total, site_MGH, site_UCLA, site_UMinn,
# site_WashU, intercept
mod94_data %>%
  select(soil_ts, age_yrs, soilxage,
         sex.numeric, household.numeric, marital.numeric,
         totalGray, rfMRI_FD_average, moca_total,
         site_MGH, site_UCLA, site_UMinn, site_WashU,
         intercept) %>%
  write.table(., file.path(abs_palm94_dir, paste0("abs_design_matrix_n", n94, ".csv")),
              row.names = FALSE, col.names = FALSE, sep = ",")

mod94_data %>%
  select(site_block) %>%
  write.table(., file.path(abs_palm94_dir, paste0("abs_site_block_n", n94, ".csv")),
              row.names = FALSE, col.names = FALSE, sep = ",")
```
